---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Patel, Nakul"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)
table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: Sex and Class are both categorical variables that describe a nominal piece of data. The remaining variables are all ratio data. WHOLE, SHUCK, RINGS, RATIO, and particularly VOLUME seem to be skewed to the right based on their mean and medians combined with the overall max value for each compared to the 3rd quartile. Of the four, VOLUME has the largest outlier and skew. The remaining variables have similar mean and medians with smaller gaps between quartiles implying less skewness and outliers.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}

t <- table(mydata$SEX, mydata$CLASS)
addmargins(t)

ggplot(data.frame(t), aes(x=Var2, y=Freq, fill=Var1, group=Var1)) + 
  geom_bar(stat = 'identity', position=position_dodge()) + 
  labs(x='CLASS', y='FREQUENCY', title= 'CLASS membership, SEX-differentiated') + 
  scale_fill_discrete(name='SEX', labels = c('Female','Infant','Male'))

```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: It is interesting that there are any infants in the A4 and A5 classes given these classes are for the oldest abalones. It is also worth noting there is a significant amount of infants through class A3 with A2 having the most. Additionally, there is a relative intra-class equivalent mix of Female and Male abalones across the various classes with Males being slightly more prevalent.***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}

set.seed(123)
sample_rows <- sample(seq(1:1036), 200)
work <- mydata[sample_rows,]
plot(work[,2:6])

```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}

ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
  geom_point(aes(color = CLASS)) + 
  labs(x = 'Volume', y = 'Whole weight', title = 'Whole weight in grams | Volume'~cm^3~'')

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}

ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + 
  geom_point(aes(color = CLASS)) + 
  labs(x = 'Whole weight', y = 'Shuck weight', title = 'Shuck weight, in grams | Whole weight, in grams') + 
  geom_abline(slope = max(mydata$SHUCK/mydata$WHOLE), intercept = 0, linetype = 2)

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: There is generally less variability in the second plot. This might imply that there is a weaker correlation between volume and whole weight than in shucked weight to whole weight. In other words, density may come into play for volume but whole weight is highly correlated to shucked meat weight. In addition, you can see in the second plot that class A5 is lower (and farther away from the maximum ratio line) than the other classes. This seems to hold true as the age class increases. This may imply that as abalones age, there is less meat weight overall compared to the whole weight.***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}

par(mfrow = c(3, 3))
hist(mydata$RATIO[mydata$SEX == 'F'], breaks = 10, col = "red", xlab = '', ylab = 'Frequency', main = 'Female RATIO')
hist(mydata$RATIO[mydata$SEX == 'I'], breaks = 10, col = "green", xlab = '', main = 'Infant RATIO')
hist(mydata$RATIO[mydata$SEX == 'M'], breaks = 10, col = "blue", xlab = '', main = 'Male RATIO')
boxplot(mydata$RATIO[mydata$SEX == 'F'], col = "red", xlab = '', main = 'Female RATIO')
boxplot(mydata$RATIO[mydata$SEX == 'I'], col = "green", xlab = '', main = 'Infant RATIO')
boxplot(mydata$RATIO[mydata$SEX == 'M'], col = "blue", xlab = '', main = 'Male RATIO')
qqnorm(mydata$RATIO[mydata$SEX == 'F'], col = "red", xlab = 'Theortical Quantities', ylab = 'Sample Quantities', main = 'Female RATIO')
qqline(mydata$RATIO[mydata$SEX == 'F'])
qqnorm(mydata$RATIO[mydata$SEX == 'I'], col = "green", xlab = 'Theortical Quantities', main = 'Infant RATIO')
qqline(mydata$RATIO[mydata$SEX == 'I'])
qqnorm(mydata$RATIO[mydata$SEX == 'M'], col = "blue", xlab = 'Theortical Quantities', main = 'Male RATIO')
qqline(mydata$RATIO[mydata$SEX == 'M'])
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: Looking across all the plots they seem to be relatively normally distributed. However, it is clear that all of the sex groups seemed to be skewed right and have outliers which confirm the original assumption about ratio being skewed right. In addition, it is clear that infant has the most oitliers from the theoretical quantities and female as the largest outliers.***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}


female_outliers <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "F"])$out, ]
infant_outliers <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "I"])$out, ]
male_outliers <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "M"])$out, ]
kable(rbind(female_outliers, infant_outliers, male_outliers))

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: The results in 3b point to a similar observation from 3a. While all of the sexes seem to have outliers, but the infant sex has the largest amount of outliers. What we can see here, that we couldn't previously is that most of the outliers are in younger classes (A1-A3) with only one in A5.***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}

grid.arrange(
  ggplot(data = mydata, aes( x=CLASS, y=VOLUME)) + geom_boxplot() + theme_bw(),
  ggplot(data = mydata, aes( x=CLASS, y=WHOLE)) + geom_boxplot() + theme_bw(),
  ggplot(data = mydata, aes( x=RINGS, y=VOLUME)) + geom_point() + theme_bw(),
  ggplot(data = mydata, aes( x=RINGS, y=WHOLE)) + geom_point() + theme_bw()
  )

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: Volume and Whole are not good predictors of age. It is generally true that the more volume or whole weight, typically the older the class. However, class A5 is the outlier here (potentially decreases in overall size due to age). Due to this it would be easy to misclassify the oldest abalone as a younger age group.***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}

aggregated_data <- aggregate(mydata[c('VOLUME', 'SHUCK', 'RATIO')], FUN=mean, by=list(mydata$SEX, mydata$CLASS))
colnames(aggregated_data) <- c('SEX', 'CLASS', 'VOLUME', 'SHUCK', 'RATIO')
kable(matrix(aggregated_data$VOLUME, nrow = 3, ncol = 5, dimnames = list(c('Female', 'Infant', 'Male'),c('A1', 'A2', 'A3', 'A4', 'A5'))), caption = 'Volume')
kable(matrix(aggregated_data$SHUCK, nrow = 3, ncol = 5, dimnames = list(c('Female', 'Infant', 'Male'),c('A1', 'A2', 'A3', 'A4', 'A5'))), caption = 'Shuck')
kable(matrix(aggregated_data$RATIO, nrow = 3, ncol = 5, dimnames = list(c('Female', 'Infant', 'Male'),c('A1', 'A2', 'A3', 'A4', 'A5'))), caption = 'Ratio')

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}

grid.arrange(
  ggplot(data = aggregated_data, aes(x = CLASS, y = RATIO, group = SEX, color = SEX)) + 
    geom_line() + labs(title = 'Mean Ratio per Class') + 
    scale_color_manual(name='SEX', labels = c('Female','Infant','Male'), values = c('red', 'green', 'blue')),
  ggplot(data = aggregated_data, aes(x = CLASS, y = VOLUME, group = SEX, color = SEX)) + 
    geom_line() + labs(title = 'Mean Volume per Class') + 
    scale_color_manual(name='SEX', labels = c('Female','Infant','Male'), values = c('red', 'green', 'blue')),
ggplot(data = aggregated_data, aes(x = CLASS, y = SHUCK, group = SEX, color = SEX)) + 
    geom_line() + labs(title = 'Mean Shuck per Class') + 
    scale_color_manual(name='SEX', labels = c('Female','Infant','Male'), values = c('red', 'green', 'blue'))
            )
```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: They seem to confirm earlier hypothesis that volume and weight, here shuck weight rather than whole weight, seem to increase with age, with class A5 seeing a slight decrease. In other words, volume and shuch weight increase until class A4 where they see a peak. A question these plots raise, however, is that females are by and large ahead in terms of mean volume and shucked weight, followed by males and then infants. It is also interesting to note that the mean ratios of females and males are similar but infant ratio is much lower except for class A1 (the youngest class). This all raises questions around when it is best to harvest abalones. Infants ate a lower stage have higher ratios which could show efficiency however class A4 is by and far the most volume and weight overall and therefore might be more realistic to harvest needed quantities to satisfy demand ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}

grid.arrange(
   ggplot(data = subset(mydata, mydata$RINGS < 16 & mydata$SEX == 'I'), aes(group=RINGS, y=VOLUME, x=RINGS)) + 
    geom_boxplot(color = 'blue') +
    labs(title = 'Infant Volume | Rings') +
    scale_x_continuous(breaks = 3:15),
  ggplot(data = subset(mydata, mydata$RINGS < 16 & (mydata$SEX == 'M' | mydata$SEX == 'F')), aes(group=RINGS, y=VOLUME, x=RINGS)) + 
    geom_boxplot(color = 'red') + 
     labs(title = 'Adult Volume | Rings') +
    scale_x_continuous(breaks = 3:15),
    ggplot(data = subset(mydata, mydata$RINGS < 16 & mydata$SEX == 'I'), aes(group=RINGS, y=WHOLE, x=RINGS)) + 
    geom_boxplot(color = 'blue') +
    labs(title = 'Infant Weight | Rings') +
    scale_x_continuous(breaks = 3:15),
  ggplot(data = subset(mydata, mydata$RINGS < 16 & (mydata$SEX == 'M' | mydata$SEX == 'F')), aes(group=RINGS, y=WHOLE, x=RINGS)) + 
    geom_boxplot(color = 'red') + 
     labs(title = 'Adult Weight | Rings') +
    scale_x_continuous(breaks = 3:15)
  
)

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: From the displays, it is clear that across the board most of the growth in terms of weight and volume for all groups occurs earlier in the life span and growth slows as abalones age. It is also clear that infants have the greatest variability in volume and weight as they get older while variability among adults is pretty similar across age groups.***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: One clear issue is the inability to properly identify sex of a abalone after breeding season leading to the misclassification of abalones as infants well into adulthood that is seen in the data. This becomes important due to the fact that we can see clear departures in the female and male abalones in terms of volume and weight with females being slightly higher. In addition, physical measurements would be more indicative if they were correlated. It is clear that there is a decrease in physical volume and weight as abalones reach the oldest age group. This makes it a non exact predictor due to this outlier behavior.***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: Not knowing about the abalone study, I would want to confirm how the population was sampled. Not knowing much about habitat or behavior, I would want to know if there was any potential bias in the sampling based on ease of sampling. I would also like to know the relative proportion of the sample size versus the overall population to understand how representative the data is.***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: Given the human nature of how this data was collected, I would question if we can pull causality due to the potential flawed nature of the data gathering. In addition, it is possible that a observational study may lead to false conclusions about causality. Without an experiment it could be easy to assume causality due to human bias (ex. size increases with age) without understandstanding of other factors that may impact the variables wanting to be studied. From these studies it is helpful to understand correlation between variables to then create hypothesizes that can be measured through experimentation. ***
